\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-02-17 doc/publications/plos16/plos16.tex
updated BMD section with PRODH as novel gene

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{GWAS-style analysis using \cursedforest\ identifies novel variants associated with Bone Mineral Density}
In this section we use \cursedforest\ to perform feature selection on over 7.2 million genomic variants to identify the
locations associated with Bone Mineral Density (BMD)~\cite{Duncan.2011}. We build a model on the BMD dataset as
described in Methods using a binary high or low BMD value as the label.  From this model, we output the top 10,000
contributing variants, ranked by their Gini impurity. The variant with the highest Gini impurity, and hence importance,
is ranked 1, the lowest 10,000.

To test whether the associated features have biological meaning, we investigate whether variants in or near
BMD-associated genes are enriched toward the top of our ranked variant list. To form the list of "known genes", we
combine the 26 known and 6 novel BMD genes listed by Duncan {\it et al.}~\cite{Duncan.2011} with genes identified by
Estrada {\it et al.}~\cite{Estrada2012, Kemp2014,Phang.2008}, resulting in 102 genes.  All genes are covered in the
10,000 extracted variants and the gene name is assigned to a variant if it falls within 100 kb of the gene.

As shown in Fig~\ref{figure:ranksumtest}, the known genes are ranked significantly higher than variants from other
locations in the genome (Mann-Whitney U, $p=9.71\times10^{-15}$), supporting our hypothesis that biologically relevant
features are contributing more information to our classification model.

The gene with the highest-ranked features is COLEC10 (collectin 10) as shown in Fig~\ref{figure:knowngeneranks}, a
well established contributor to BDM~\cite{Kemp2014,Liu2008} with verified transcription factor binding site for FOXL1,
also a known BDM gene~\cite{DECODE}.  The second highest ranked gene is ZBTB40 on 22q11. Stagi {\it et
  al.}~\cite{Stagi.2010}, observe that individuals with a micro-deletion in this region have a significant reduction in
bone mass. In this same region, we also observe the next highest ranked gene, PRODH (proline dehydrogenase).  PRODH
catalyses the first step in proline degradation, and as proposed by Phang {\it et al.}~\cite{Phang.2008}, proline is
made available by collagen degradation.

The lowest ranked gene with literature implied BMD association is DCDC5 (Doublecortin Domain Containing 5) ranked 9,830
out of 10,000 (see Fig~\ref{figure:knowngeneranks}).  Similar to the conclusions drawn by Duncan {\it et
  al.}~\cite{Duncan.2011}, our results also indicate that this gene does not have a strong association with BMD.  In
fact, DCDC5 is not highly expressed in bone, and any mechanisms by which this gene might regulate BMD are unclear
\cite{Thakker2012}.  Although variants near DCDC5 are associated with BMD \cite{Rivadeneira2009}, it is questionable as
to whether the gene itself contributes to BMD.

Hence focusing on the top 20 most associated variants, the majority are from five known BMD genes (left side of the
vertical line in Fig~\ref{figure:knowngeneranks}), however there are also five novel locations as annotated in
Fig~\ref{figure:ranksumtest}. Most notably the SNP in the intronic region of Unc13c (rs12916554) is linked to quality
of sleep~\cite{Nicod2016}, which in turn is reported to be the best predictor of osteoporosis risk~\cite{Magali2016}, a
condition diagnosed by measuring BMD.  Also listed is rs1035665, which was identified in a recent GWAS study to regulate
ABCF1 (ATP Binding Cassette Subfamily F Member 1)~\cite{Dixon2007} with member 2 of this subfamily being a known BMD
gene. The other three novel loci are on or near ncRNA.

@@

\begin{figure}[tbhp]
  \begin{adjustwidth}{-1.00in}{0in}
    \caption{\textbf{Distribution of known BMD genes amongst the top 10,000 features}}
    \label{figure:ranksumtest}
    \includegraphics[totalheight=8cm]{./figs/BMDTop10K.pdf}
    \begin{flushleft}
      The figure shows the top 10,000 features by their rank and Gini impurity score (Importance). The line denotes the top 20 features which contains five novel loci annotated with text. The $p$-value is the Mann-Whitney U test.
     \end{flushleft}
  \end{adjustwidth}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{GWAS-style analysis using \cursedforest\ identifies novel variants associated with Bone Mineral Density}
In this section we use \cursedforest\ to perform feature selection on over 7.2 million genomic variants and identify the
locations associated with Bone Mineral Density (BMD)~\cite{Duncan.2011}. 
%We build a model on the BMD dataset as described in Methods using a binary high or low BMD value as the label.  
We rank variants by their Gini impurity, with the most important one ranked at position 1, and return the top 10,000 variants.

To test whether the associated variants have biological meaning, we investigate whether variants in or near
BMD-associated genes are enriched toward the top of our ranked list. To form the list of "known genes", we
combine the 26 known and 6 novel BMD genes listed by Duncan {\it et al.}~\cite{Duncan.2011} with the genes identified by
Estrada {\it et al.}~\cite{Estrada2012} and Lui {\it et al.}~\cite{Liu2008}, resulting in 102 genes.  All genes are covered in the
10,000 extracted variants and the gene name is assigned to a variant if it falls within 100 kb of the gene.

As shown in Fig~\ref{figure:ranksumtest}, the known genes are ranked significantly higher than variants from other
locations in the genome (Mann-Whitney U, $p=1.83\times10^{-14}$), supporting our hypothesis that biologically relevant
features are contributing more information to our classification model.

The two gene with the highest-ranked features are the known BMD associated genes COLEC10 (collectin 10) and ZBTB40 (Zinc Finger And BTB Domain-Containing Protein 40) as shown in Fig~\ref{figure:knowngeneranks}. 
The association for both gene has been replicated since in independent studies (COLEC10~\cite{Kemp2014} and ZBTB40~\cite{Nielson2016}).
%with verified transcription factor binding site for FOXL1,
%also a known BDM gene~\cite{DECODE}.  
%The second highest ranked gene is ZBTB40, also a well established BMD associated gene~\cite{27476799}.

In contrast, the lowest ranked gene with literature implied BMD association is DCDC5 (Doublecortin Domain Containing 5) ranked 9,830
out of 10,000 (see Fig~\ref{figure:knowngeneranks}).  Similar to the conclusions drawn by Duncan {\it et
  al.}~\cite{Duncan.2011}, our results also indicate that this gene does not have a strong association with BMD.  In
fact, DCDC5 is not highly expressed in bone, and any mechanisms by which this gene might regulate BMD are unclear
\cite{Thakker2012}.  Although variants near DCDC5 are associated with BMD \cite{Rivadeneira2009}, it is questionable as
to whether the gene itself contributes to BMD.

Hence focusing on the top 20 most associated variants, the majority are from the four known BMD genes (left side of the
vertical line in Fig~\ref{figure:knowngeneranks}), however there are also six novel locations as annotated in
Fig~\ref{figure:ranksumtest}. Most notably PRODH (proline dehydrogenase) located on 22q11, a loci observed by Stagi {\it et
  al.}~\cite{Stagi.2010} to significantly reduce bone mass in individuals with a micro-deletion in this region. The variant in PRODH is the only 
  highly associated feature in this region and not in linkage disequilibrium (LD) with another gene (see Fig~\ref{zoomplot}). PRODH
catalyses the first step in proline degradation, and as proposed by Phang {\it et al.}~\cite{Phang.2008}, proline is
made available by collagen degradation. Also amongst the top 20 associated variants is a SNP in the intronic region of Unc13c (rs12916554), a gene that is linked to quality
of sleep~\cite{Nicod2016}, which in turn is reported to be the best predictor of osteoporosis risk~\cite{Magali2016}, a
condition diagnosed by measuring BMD.  Also listed is rs1035665, which was identified in a recent GWAS study to regulate
ABCF1 (ATP Binding Cassette Subfamily F Member 1)~\cite{Dixon2007} with member 2 of this subfamily being a known BMD
gene. The other three novel loci are on or near ncRNA.

@@

\begin{figure}[tbhp]
  \begin{adjustwidth}{-2.00in}{0in}
    \caption{\textbf{Importance of 10,000 top ranked features and visualisation of the 22q.11 region}}
    \label{figure:ranksumtest}
     \begin{subfigure}[b]{0.5\linewidth}
    \includegraphics[totalheight=7cm]{./figs/BMDTop10K.pdf}
     \caption{Distribution of known BMD genes amongst the top 10,000 features.} 
      \label{figure:rbo-prod.png} 
%      \vspace{4ex}
    \end{subfigure} 
 \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=7cm]{./figs/EURchr22_18-20mb.pdf}
      \caption{Zoom plot of the 22q.11 region centred on PRODH.} 
      \label{zoomplot} 
%      \vspace{4ex}
    \end{subfigure} 
    \begin{flushleft}
      (a) The figure shows the top 10,000 features by their rank and Gini impurity score (Importance). The line denotes the top 20 features which contains five novel loci annotated with text. The $p$-value is the Mann-Whitney U test.
      (b) Each point represents a variant with the normalised Gini impurity score on the left y axis and LD in colour with the highest $r^2$ in red.  
     \end{flushleft}
  \end{adjustwidth}
\end{figure}
